---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.path = 'figure/')
```

## Loading and preprocessing the data
To perform this analysis, one need to fork/clone the data provided on this [github repository](http://github.com/rdpeng/RepData_PeerAssessment1).

The zip file "activity.zip" contains a single file activity.csv. Validate this by using the following command to inspect the contents of a zip file:
```{r}
filename = "activity.zip"
unzip(filename, list = TRUE)
```

Read the data using read.csv command:
```{r message = FALSE}
data <- read.csv(unz(filename, "activity.csv"))
summary(data)
dim(data)
# importing dpylr, maybe useful for some data minpulation
library(dplyr)
```
## What is mean total number of steps taken per day?

Calculate total number of steps taken each day, ignoring NA values:
```{r}
dailytotal <- data %>%
     na.omit() %>%
     group_by(date) %>%
     summarise(total = sum(steps))
head(dailytotal)
```

Now, show a histogram of total number of steps taken each day:
```{r}
hist(dailytotal$total, breaks = 30, xlab = NULL, main = "Total number of steps taken each day")
```

Display mean and median of the total number of steps taken each day:
```{r}
summary(dailytotal$total)
```
**ANSWER** *Mean is 10,770 steps and median is 10,760 for total steps taken each day*

## What is the average daily activity pattern?
Calculate average number of steps taken per interval, across all days
```{r}
dailyactivity <- data %>%
        na.omit() %>%
        group_by(interval) %>%
        summarise(mean = mean(steps))
```

Plot average daily activility pattern
```{r}
with(dailyactivity, plot(interval, mean, type = "l", xlab = "Interval", ylab = "Average number of steps taken", main = "Average daily activity pattern"))
```

Interval 835 contains the maximum number of steps, on average across all days in the dataset. In real terms, this would be around 1:55pm of the day.
```{r}
with(dailyactivity, dailyactivity[which(mean == max(mean)),])
```

**ANSWER** *Interval 835 contains the maximum number of steps.*

## Imputing missing values
There are 2304 rows with NA out of 17568 rows. The only column containing NA is "steps".
```{r}
sum(!complete.cases(data))
```

After inspecting the NA data, a date is either entirely consisted of NA entries or consisted of valid observations. Here is a table showing number of NA by day:
```{r}
head(table(data$date, is.na(data$steps)),20)
```

The strategy I propose would be to use median number of steps for each interval, across all days to fill the NA. First, prepare a table with median step by interval, across all days.
```{r}
medianbyinterval <- data %>%
        na.omit() %>%
        group_by(interval) %>%
        summarise(median = median(steps))
```

Now, we assign the NA with the imputed values. We merge the median table and assign steps based on a if condition. We will use mapply for this step. The new data table is called *data.new*:
```{r}
mergedtbl <- merge(data, medianbyinterval, by = "interval")
mergedtbl <- mutate(mergedtbl, steps = mapply(ifelse, is.na(steps), median, steps))
data.new <- select(mergedtbl, steps, date, interval)
```


Let's look at histogram for both new and original data:
```{r}
dailytotal.new <- data.new %>%
     na.omit() %>%
     group_by(date) %>%
     summarise(total = sum(steps))
par(mfrow = c(2,1), oma = c(2,2,0,0), mar = c(1,1,3,3))
hist(dailytotal$total, breaks = 30, xlab = NULL, main = "Original")
hist(dailytotal.new$total, breaks = 30, xlab = NULL, main = "New")
```

It turns out, the imputation added values in the low 1000 range. One observes a boost in the 2nd column. Now, we will look at the mean and median to see if there is any difference:
```{r}
print("Original data")
summary(dailytotal$total)
print("New data after imputation")
summary(dailytotal.new$total)
```
The mean and median are lower for the new dataset compared to the original.Let's also look at an boxplot comparison.
```{r}
new <- data.frame(total = dailytotal.new$total, type = "new")
original <- data.frame(total = dailytotal$total, type = "original")
combined <- rbind.data.frame(new,original)
boxplot(combined$total ~ combined$type, main = "Total number of steps taken each day")
points(1:2, c(mean(dailytotal.new$total), mean(dailytotal$total)), pch = 16, col = 2)
legend("topright", legend = "mean", pch = 16, col = 2)
```
**Strategy** *I decided to imput using median steps for each interval, across all days that have values.*
## Are there differences in activity patterns between weekdays and weekends?

First, we created a new variable called weekend in the data table. A helper function is created to help with this.

```{r}
whatday <- function(x) {
        day <- weekdays(strptime(x, "%Y-%m-%d"))
        ifelse(day %in% c("Sunday", "Saturday"), "weekend", "weekday")
}
data.new$weekend <- whatday(data.new$date)
```

Now, we are going to plot the data of average number of steps taken, averaged across all weekday days or weekend days.
```{r}
library(lattice)
mean <- data.new %>%
        na.omit() %>%
        group_by(interval, weekend) %>%
        summarise(mean = mean(steps))
xyplot(mean ~ interval | weekend, mean, type = "l", layout = c(1,2), xlab = "Interval", ylab = "Number of Steps")
```

**ANSWER** *The data indicates the subject wakes up earlier and peak activity is more drastic during weekdays. During weekend, subject stays up later compared to weekdays.*